\pkg{dataMaid}: your assistant for data cleaning in \proglang{R}

### 2017-01-24 latex/article_vol2.tex
Bibliography

As with \R{summaryFunction}s, a complete list of available
\R{checkFunction}s is obtained by calling
\R{allCheckFunctions()}. Note however, that \R{checkFunction}s have a
usage beyond the \R{XXXChecks} arguments, namely in the
\textit{pre-check} stage. In this stage, it is determined whether or
not each variable is suitable for the summarize/visualize/check (SVC)
steps. The functions used in the pre-check stage should be
\R{checkFunction}s that are applicable to all variable classes. The
default pre-checks, the functions \R{isKey()} and \R{isEmpty()}, check
whether a variable has unique values for all observations or only a
single value for all observations, respectively. If one of these
statements are true, the variable will not be subjected to the SVC
steps.  We can allow empty variables to move on to the SVC step by
only checking for keys in the pre-check step:

\begin{Schunk}
\begin{Sinput}
> clean(toyData, preChecks = "isKey")
\end{Sinput}
\end{Schunk}

Note that the data visualizations in the cleaning summary are also
controllable, though only a single function can be provided for all
variable types. If for instance, we wish to change the visualizations
from the default \pkg{ggplot2} style histograms and barplots to base
\proglang{R} histograms and barplots, we type
\begin{Schunk}
\begin{Sinput}
> clean(toyData, allVisuals = "basicVisual")
\end{Sinput}
\end{Schunk}

%\hl{
%\begin{itemize}
%\item Introduce the relevant \R{clean}-arguments and the \R{defaultWhateverSummaries} etc.-functions
%\item Introduce \R{allSummaryFunctions()} etc. and present a table corresponding to the output of this call
%\item Small examples:
%\begin{itemize}
%\item Add a function to one of the XXXSummaries/XXXVisuals/XXXChecks-arguments (still calling default options)
%\item Remove all but a single function from one of these arguments
%\item Describe what happens if the argument is NULL
%\end{itemize}
%\end{itemize}
%}

In summary, and as indicated in Figure~\ref{figure:cleanStructure}, there are two stages
where \R{clean()} applies functions to each of the variables:
\begin{enumerate}
\item In the pre-check stage
\item As part of the summarize/visualize/check (SVC) steps
\end{enumerate}
Each of these stages are controllable using appropriate function
arguments in \R{clean()}, and above we have shown examples of how to
tweak them to modify the data cleaning outputs. However, if for
instance the dataset at hand requires new, additional checks, then more control is needed, and
Section~\ref{sec:extending} explains how to modify and expand the
possibilities by producing new summary-, visual- and check functions.

\section{Using dataMaid interactively}
\label{sec:interactiveCleanR}

While overview documents are great for presenting and documenting the
data cleaning checks, it may be useful to be able to work more interactively
through the data cleaning process as well. Aside from the \R{clean()} function
presented above, \pkg{dataMaid} also provides more standard \proglang{R} interactive tools, such as
functions that print results to the console or returns the information
as an object for later use. This section describes how to use the
functions \R{check()}, \R{summarize()} and \R{visualize()} to work
interactively with \pkg{dataMaid}.

\subsection{Data cleaning by hand: An example}
Let's say we wish to look further into a certain variable from
\R{toyData}, namely \R{var2}. The data cleaning summary found some
issues in this variable, and we would like to recall what these issues
were. This can be done using the \R{check()} command
\begin{Schunk}
\begin{Sinput}
> check(toyData$var2)
\end{Sinput}
\begin{Soutput}
$identifyMissing
The following suspected missing value codes enter as regular values: 999, NaN.
$identifyOutliers
Note that the following possible outlier values were detected: 82, 999.
\end{Soutput}
\end{Schunk}
% $

Note that the arguments specifying which checks to perform, as
described in the previous section, are in fact passed to \R{check()},
and thus they can also be used here. For instance, if we only want to
check for potentially miscoded missing values, we can use the relevant
\R{XXXChecks} argument (e.g., \R{numericChecks}, \R{factorChecks},
etc. as described in Section~\ref{sec:usingdataMaid}) to provide a
vector of the check functions that should be applied. Recall that
Table~\ref{table.SVCfunctions} or an \R{allCheckFunctions()} call provide
overviews of the available check functions.
Moving forward, we limit the numeric checks to only identify miscoded
missing values:

\begin{Schunk}
\begin{Sinput}
> check(toyData$var2, numericChecks = "identifyMissing")
\end{Sinput}
\begin{Soutput}
$identifyMissing
The following suspected missing value codes enter as regular values: 999, NaN.
\end{Soutput}
\end{Schunk}

An equivalent way to call only a single, specific \R{checkFunction},
such as \R{identifyMissing}, is by using it directly on the variable, e.g.,
\begin{Schunk}
\begin{Sinput}
> identifyMissing(toyData$var2)
\end{Sinput}
\begin{Soutput}
The following suspected missing value codes enter as regular values: 999, NaN.
\end{Soutput}
\end{Schunk}
% $

The result of a \R{checkFunction} is an object of class
\R{checkResult}. By using the structure function, \R{str()}, we can
look further into its components:

\begin{Schunk}
\begin{Sinput}
> missVar2 <- identifyMissing(toyData$var2)
> str(missVar2)
\end{Sinput}
\begin{Soutput}
List of 3
 $ problem      : logi TRUE
 $ message      : chr "The following suspected missing value codes enter as
 		 regular values: \\\"999\\\", \\\"NaN\\\"."
 $ problemValues: num [1:2] 999 NaN
 - attr(*, "class")= chr "checkResult"
\end{Soutput}
\end{Schunk}
The most important thing to note here is that while the printed
message is made for easy reading, the actual values of the variable
causing the issue are still obtainable in the element
\R{problemValues}. If we for instance decide that the values \R{999}
and \R{NaN} in \R{var2} are in fact miscoded missing values, we can
easily replace them with \R{NA}s:
\begin{Schunk}
\begin{Sinput}
> toyData$var2[toyData$var2 %in% missVar2$problemValues] <- NA
> identifyMissing(toyData$var2)
\end{Sinput}
\begin{Soutput}
No problems found.
\end{Soutput}
\end{Schunk}
% $

Similarly, the \R{visualize()} and \R{summarize()} functions can be
used to run the corresponding visualizations and summaries for each
variable. See Figure~\ref{fig:example2} for the visualization output.

\begin{Schunk}
\begin{Sinput}
> visualize(toyData$var2)
> summarize(toyData$var2)
\end{Sinput}
\begin{Soutput}
     Feature                   Result
[1,] "Variable type"           "numeric"
[2,] "Number of missing obs."  "3 (20 %)"
[3,] "Number of unique values" "8"
[4,] "Median"                  "4.5"
[5,] "1st and 3rd quartiles"   "1.75; 6"
[6,] "Min. and max."           "1; 999"
\end{Soutput}
\end{Schunk}
% $

\begin{figure}[tb]
\begin{center}
\includegraphics[width=7.5cm]{toyData-var2.pdf}
\end{center}
\label{fig:example2}
\caption{Output from running \R{visualize()} on the variable \texttt{var2} from the
\R{toyData} dataset.}
\end{figure}

As we saw with the \R{check()} function, the summary can be modified
by providing the relevant \R{XXXSummaries} arguments. Setting the
\R{numericSummaries} argument, we can control the summary output by
providing a vector of function names to run for a particular summary. To
only get the median, minimum and the maximum we set
\R{numericSummaries=c("centralValue", "minMax")}:

\begin{Schunk}
\begin{Sinput}
> summarize(toyData$var2, numericSummaries = c("centralValue", "minMax"))
\end{Sinput}
\begin{Soutput}
     Feature         Result
[1,] "Median"        "4.5"
[2,] "Min. and max." "1; 999"
\end{Soutput}
\end{Schunk}
% $

Note that the \R{summarize()}, \R{check()} and \R{visualize()} functions are also available interactively for full datasets, by calling e.g., \R{summarize(toyData)}. However, this produces an extensive amount of output in the console, and therefore, we generally do not recommend it, unless working with very small datasets or subsets of datasets.

@@

\section{Extending dataMaid}
\label{sec:extending}

\pkg{dataMaid} can be used as a user-friendly, self-contained package, as
shown in the previous sections. However, \pkg{dataMaid} is fully
customizable and \R{clean()} is mainly a tool for formatting the
results from various checking-, summary- and visualization
functions. Thus, the actual work underlying a \pkg{dataMaid} output file
can be anything --- depending on the arguments given to \R{clean()}
--- and user made functions are easily added to the
summarize/visualize/check-function arguments, as discussed previously.
However, the functions used in the SVC steps must adhere
to specific structures in order to be called from these three steps, and
therefore, we will now present how \R{summaryFunction}s,
\R{visualFunction}s and \R{checkFunction}s are made.

This section consists of two parts. First, we describe how customized \R{summaryFunction}s,
\R{visualFunction}s and \R{checkFunctions} can be made, including an introduction to a
few convenient tools available in \pkg{dataMaid} for aiding SVC-function construction.
Table \ref{table.functionTypes} serves as an overview of the internals of these three
central function types. Secondly, we turn to a
worked example of how to use custom made functions in practice. Here,
four new SVC functions are defined and used, both interactively and in \R{clean()}.

\begin{table}[htbp]
\footnotesize
\bgroup
\def\arraystretch{1.8}%
\begin{tabular}{L{0.2\linewidth}L{0.2\linewidth}L{0.2\linewidth}L{0.2\linewidth}}
& \R{summaryFunction} & \R{visualFunction} & \R{checkFunction} \\
\hline
\vspace{0pt} Input (required) & \R{v} - a variable vector \newline \R{...}  &  \R{v} - a variable vector \newline \R{vnam} - the variable name (as character string) \newline \R{doEval} - a logical (\R{TRUE}/\R{FALSE}) controlling the output type of the function & \R{v} - a variable vector \newline \R{nMax} - an integer (or \R{Inf}), controlling how many problematic values are printed, if relevant \newline \R{...}  \\
Input (optional) &  \R{maxDecimals} - number of decimals printed in outputted numerical values.  & - &  \R{maxDecimals}  - number of decimals printed in outputted numerical values.  \\
Purpose & Describe some aspect of the variable, e.g., a central value, its dispersion or level of missingness. & Produce a distribution plot. & Check a variable for a specific issue and, if relevant, identify the values in the variable that cause the issue. \\
Output (required) & A list with entries: \newline \R{\$feature} - a label for the summary value (as character string); \newline \R{\$result} - the result of the summary (as character string) & A character string with \R{R} code for producing a plot. This code should be standalone, i.e., should include the data if necessary. & A list with entries: \newline \R{\$problem} - a logical identifying whether an issue was found; \newline \R{\$message} - a character string (possibly empty) decribing the issue that was found, properly escaped and ready for use in \R{rmarkdown} \\
Output (recommended) & A \R{summaryResult} object, i.e., an attributed list with entries \R{\$feature}, \R{\$result} and \R{\$value}, the latter being the values from \R{\$result} in their original format). & \bgroup
\def\arraystretch{1.8}%
\hspace*{-.2cm}\begin{tabular}[t]{p{0.45\linewidth} |
                 p{0.45\linewidth}} \vspace*{-1cm}\raggedright If
                 \R{doEval} is \R{TRUE}: \newline A plot that is
                 opened by the graphic device in \R{R}. &
                                                          \vspace*{-1cm}\raggedright If \R{doEval} is \R{FALSE}: \newline   A text string with \R{R} code, as described above. \end{tabular} \egroup  & A \R{checkResult} object, i.e., an attributed list with entries \R{\$problem}, \R{\$message} and \R{\$problemValues}, the latter being either \R{NULL} or the problem causing values, as they were found in \R{v}, whichever is relevant. \\
Tools available for producing the function & \R{summaryResult()} & --- & \R{messageGenerator()} \newline \R{checkResult()}
\end{tabular}
\egroup
\caption{Reference information for creating new functions to be used
  as part of the \mbox{summarize-,} visualize-, and check steps. The three
  columns correspond to each of the three function types. %\hl{I think the formatting here is still sort of chaotic. Maybe include horizontal lines everywhere? Is it allowed in JSS? It's impossible to read what's in which row right now.}
}
\label{table.functionTypes}
\end{table}

\subsection{Function templates}
\label{sec:functionTemplates}
In order to construct a summary-, visual or check function, one needs
to create a new function with a specific structure.  This can be done with
different levels of strictness. If
the new custom function is only to be used as part of the SVC steps in \R{clean()},
then only the input/output structure of the function needs special attention.
However, new user-defined functions can also be registered locally to
be part of the full machinery of \pkg{dataMaid}, and these function will
be recognized and behave in the same way as the build-in functions in
\pkg{dataMaid}. The presentation below is given in the format of
function templates, written in pseudo-code. These templates are
designed for getting the full functionality, but note that
Table~\ref{table.functionTypes} serves as a reference to the minimal
requirements, while also presenting the "full" versions of the
function types.

@@

\subsubsection{Writing a summaryFunction}
As mentioned above, \pkg{dataMaid} provides a dedicated class for
\code{summaryFunction}s. However, this does not imply that they are
particularly advanced or complicated to create; in fact, they are
nothing but regular functions with a particular
input/output-structure. Specifically, they all follow the template
below:
\begin{Verbatim}
mySummaryFunction <- function(v, ...) {
  res <- [result of whatever summary we are doing]
  summaryResult(list(feature = "[Feature name]", result = res))
}
\end{Verbatim}
The last function called here, \R{summaryResult()}, changes the class
of the output, thereby making a \R{print()} method available for it.
Note that \code{v} is a vector and that \code{res} should be either a
character string or something that will be printed as one. In other
words, e.g., integers are allowed, but matrices are not. Though a lot
of different things can go into the \code{summaryFunction} template,
we recommend only using it for summarizing the features of a variable,
and leaving tests and checks for the \code{checkFunction}s (presented
below).

Adhering to the template above is sufficient for using the freshly
made \R{mySummaryFunction()} in \R{clean()}, but we recommend
furthermore adding the new function to the overview of all summary
functions by converting it to a proper \code{summaryFunction}
object. This is done by calling the \R{summaryFunction()} creator with
the user-defined function as the first argument, and additional arguments
\R{description} (an explanatory text which will be added to the attributes of
the function), and \R{classes} (a vector of variable classes the
user-defined function is intended to be applied to, also stored as an
attribute). In other words, a call on the following form should be made:
\begin{Schunk}
\begin{Sinput}
mySummaryFunction <- summaryFunction(mySummaryFunction,
  description = "[Text describing what the summaryFunction does]",
  classes = c([vector of data types that the function is intended for]))
\end{Sinput}
\end{Schunk}
which adds the new function to the output of an
\code{allSummaryFunctions()} call.
% One comment should be devoted to
% the two attributes of a \R{summaryFunction}.
If the \R{description} argument is left unspecified, the name of the
function (which in this case is \R{"mySummaryFunction"}) will be filled in and
stored under the \R{description} attribute. What happens if the \R{classes} argument is
not specified depends on the type of \R{mySummaryFunction}. If
\R{mySummaryFunction} is constructed as an \R{S3} generic function with associated
methods, the call to \R{summaryFunction()} will automatically produce
a vector of the names of the classes for which the function can be
called. If \R{mySummaryFunction} is not an \R{S3} generic and
\R{classes} is left unspecified, the attribute will simply be
left empty. Note that the helper function \R{allClasses()} might be useful
for filling out the \R{classes} argument, as it simply lists all
available classes in \pkg{dataMaid}:
\begin{Schunk}
\begin{Sinput}
> allClasses()
\end{Sinput}
\begin{Soutput}
[1] "character" "Date"      "factor"   "integer"   "labelled"
[6] "logical"  "numeric"
\end{Soutput}
\end{Schunk}
% \hl{Write something here, don't end paragraph with code. Also, maybe move the allClasses() stuff somewhere else, it doesn't really belong under this header. Not sure where to, though.}

%Jeg foreslår at droppe nedenstående, da jeg synes det bliver lidt søgt i forhold
%til at det kun er beskrevet i pseudokode. (fx. er classes = numeric meget arbitrært)
%If we run the \R{allSummaryFunctions()} function we can see that the
%new function has been registered (output slightly appreviated):

@@

\subsubsection{Writing a visualFunction}
\code{visualFunction}s are the functions that produce the figures in a
\pkg{dataMaid} output document. Writing a \R{visualFunction} is slightly
more complicated than writing a \R{summaryFunction}. This follows from
the fact that \R{visualFunctions} need to be able to output standalone
code for plots in order for \code{clean()} to build standalone
\pkg{rmarkdown} files. We recommend using the following structure:
\begin{Schunk}
\begin{Sinput}
myVisualFunction <- function(v, vnam, doEval) {
  thisCall <- call("[the name of the function used to produce the plot]",
    v, [additional arguments for the plotting function])
  if (doEval) {
    return(eval(thisCall))
  } else return(deparse(thisCall))
}
\end{Sinput}
\end{Schunk}

In this function, \code{v} is the variable to be visualized,
\code{vnam} is its name (which should generally be passed to
\code{title} or \code{main} arguments in plotting functions) and
\code{doEval} controls whether the output is a plot (if \code{TRUE})
or a character string of standalone code for producing a plot (if
\code{FALSE}). Implementing the \code{doEval = TRUE} setting is not
strictly necessary for a \R{visualFunction}'s use in \code{clean()}, but
it makes it easier to assess what visualization options are available,
and obviously, it is crucial for interactive usage of
\R{myVisualFunction()}. In either case, it should be noted that all
the parameters listed above, \code{v}, \code{vnam} and \code{doEval},
are mandatory, so they must be left as is, even if they are not in
use (c.f. Table~\ref{table.functionTypes}).

As with the summary function, we call \R{visualFunction()} to
register our newly created function:
%. Like before, it accepts the
%user-defined function as the first argument, and two additional
%arguments description which adds explanatory flavour text as an
%attribute of the function, and a vector which determines which classes
%the user-defined function should be applied to (which is also stored
%as an attribute).
\begin{Schunk}
\begin{Sinput}
myVisualFunction <- visualFunction(myVisualFunction,
  description = "[Some text describing the visualFunction]",
  classes = c([data types that this function is intended for]))
  )
\end{Sinput}
\end{Schunk}

Now, \R{myVisualFunction()} will be available in a \R{allVisualFunctions()}
call, just like the two build-in \R{visualFunction}s, \R{standardVisual}
and \R{basicVisual}.

>>>>

As with \code{summaryFunction}s, a complete list of available
\code{checkFunction}s is obtained by calling
\code{allCheckFunctions()}. Note however, that \code{checkFunction}s have a
usage beyond the \code{XXXChecks} arguments, namely in the
\textit{pre-check} stage. In this stage, it is determined whether or
not each variable is suitable for the summarize/visualize/check (SVC)
steps. The functions used in the pre-check stage should be
\code{checkFunction}s that are applicable to all variable classes. The
default pre-checks, the functions \code{isKey()} and \code{isEmpty()}, check
whether a variable has unique values for all observations or only a
single value for all observations, respectively. If one of these
statements are true, the variable will not be subjected to the SVC
steps.  We can allow empty variables to move on to the SVC step by
only checking for keys in the pre-check step:

\begin{Schunk}
\begin{Sinput}
R> clean(toyData, preChecks = "isKey")
\end{Sinput}
\end{Schunk}

Note that the data visualizations in the cleaning summary are also
controllable, though only a single function can be provided for all
variable types. If for instance, we wish to change the visualizations
from the default \pkg{ggplot2} style histograms and barplots to base
\proglang{R} histograms and barplots, we type

\begin{Schunk}
\begin{Sinput}
R> clean(toyData, allVisuals = "basicVisual")
\end{Sinput}
\end{Schunk}

%\hl{
%\begin{itemize}
%\item Introduce the relevant \code{clean}-arguments and the \code{defaultWhateverSummaries} etc.-functions
%\item Introduce \code{allSummaryFunctions()} etc. and present a table corresponding to the output of this call
%\item Small examples:
%\begin{itemize}
%\item Add a function to one of the XXXSummaries/XXXVisuals/XXXChecks-arguments (still calling default options)
%\item Remove all but a single function from one of these arguments
%\item Describe what happens if the argument is NULL
%\end{itemize}
%\end{itemize}
%}

In summary, and as indicated in Figure~\ref{figure:cleanStructure}, there are two stages
where \code{clean()} applies functions to each of the variables:
\begin{enumerate}
\item In the pre-check stage
\item As part of the summarize/visualize/check (SVC) steps
\end{enumerate}
Each of these stages are controllable using appropriate function
arguments in \code{clean()}, and above we have shown examples of how to
tweak them to modify the data cleaning outputs. However, if for
instance the dataset at hand requires new, additional checks, then more control is needed, and
Section~\ref{sec:extending} explains how to modify and expand the
possibilities by producing new summary-, visual- and check functions.

\section[Using dataMaid interactively]{Using \pkg{dataMaid} interactively}
\label{sec:interactiveCleanR}

While overview documents are great for presenting and documenting the
data cleaning checks, it may be useful to be able to work more interactively
through the data cleaning process as well. Aside from the \code{clean()} function
presented above, \pkg{dataMaid} also provides more standard \proglang{R} interactive tools, such as
functions that print results to the console or returns the information
as an object for later use. This section describes how to use the
functions \code{check()}, \code{summarize()} and \code{visualize()} to work
interactively with \pkg{dataMaid}.

\subsection{Data cleaning by hand: An example}
Let's say we wish to look further into a certain variable from
\code{toyData}, namely \code{var2}. The data cleaning summary found some
issues in this variable, and we would like to recall what these issues
were. This can be done using the \code{check()} command

\begin{Schunk}
\begin{Sinput}
R> check(toyData$var2)
\end{Sinput}
\begin{Soutput}
$identifyMissing
The following suspected missing value codes enter as regular values: 999, NaN.
$identifyOutliers
Note that the following possible outlier values were detected: 82, 999.
\end{Soutput}
\end{Schunk}

Note that the arguments specifying which checks to perform, as
described in the previous section, are in fact passed to \code{check()},
and thus they can also be used here. For instance, if we only want to
check for potentially miscoded missing values, we can use the relevant
\code{XXXChecks} argument (e.g., \code{numericChecks}, \code{factorChecks},
etc. as described in Section~\ref{sec:usingdataMaid}) to provide a
vector of the check functions that should be applied. Recall that
Table~\ref{table.SVCfunctions} or an \code{allCheckFunctions()} call provide
overviews of the available check functions.
Moving forward, we limit the numeric checks to only identify miscoded
missing values:

\begin{Schunk}
\begin{Sinput}
R> check(toyData$var2, numericChecks = "identifyMissing")
\end{Sinput}
\begin{Soutput}
$identifyMissing
The following suspected missing value codes enter as regular values: 999, NaN.
\end{Soutput}
\end{Schunk}

An equivalent way to call only a single, specific \code{checkFunction},
such as \code{identifyMissing}, is by using it directly on the variable,
e.g.,

\begin{Schunk}
\begin{Sinput}
R> identifyMissing(toyData$var2)
\end{Sinput}
\begin{Soutput}
The following suspected missing value codes enter as regular values: 999, NaN.
\end{Soutput}
\end{Schunk}

The result of a \code{checkFunction} is an object of class
\code{checkResult}. By using the structure function, \code{str()}, we can
look further into its components:

\begin{Schunk}
\begin{Sinput}
R> missVar2 <- identifyMissing(toyData$var2)
R> str(missVar2)
\end{Sinput}
\begin{Soutput}
List of 3
 $ problem      : logi TRUE
 $ message      : chr "The following suspected missing value codes enter as regular values: \\\"999\\\", \\\"NaN\\\"."
 $ problemValues: NULL
 - attr(*, "class")= chr "checkResult"
\end{Soutput}
\end{Schunk}

The most important thing to note here is that while the printed
message is made for easy reading, the actual values of the variable
causing the issue are still obtainable in the element
\code{problemValues}. If we for instance decide that the values \code{999}
and \code{NaN} in \code{var2} are in fact miscoded missing values, we can
easily replace them with \code{NA}s:

\begin{Schunk}
\begin{Sinput}
R> toyData$var2[toyData$var2 %in% missVar2$problemValues] <- NA
R> identifyMissing(toyData$var2)
\end{Sinput}
\begin{Soutput}
The following suspected missing value codes enter as regular values: 999, NaN.
\end{Soutput}
\end{Schunk}

Similarly, the \code{visualize()} and \code{summarize()} functions can be
used to run the corresponding visualizations and summaries for each
variable. See Figure~\ref{fig:example2} for the visualization output.

\begin{Schunk}
\begin{Sinput}
R> visualize(toyData$var2)
R> summarize(toyData$var2)
\end{Sinput}
\begin{Soutput}
     Feature                   Result    
[1,] "Variable type"           "numeric" 
[2,] "Number of missing obs."  "3 (20 %)"
[3,] "Number of unique values" "8"       
[4,] "Median"                  "4.5"     
[5,] "1st and 3rd quartiles"   "1.75; 6" 
[6,] "Min. and max."           "1; 999"  
\end{Soutput}
\end{Schunk}

\begin{figure}[tb]
\begin{center}
\includegraphics[width=7.5cm]{toyData-var2.pdf}
\end{center}
\label{fig:example2}
\caption{Output from running \code{visualize()} on the variable \texttt{var2} from the
\code{toyData} dataset.}
\end{figure}

As we saw with the \code{check()} function, the summary can be modified
by providing the relevant \code{XXXSummaries} arguments. Setting the
\code{numericSummaries} argument, we can control the summary output by
providing a vector of function names to run for a particular summary. To
only get the median, minimum and the maximum we set
\code{numericSummaries=c("centralValue", "minMax")}:

\begin{Schunk}
\begin{Sinput}
R> summarize(toyData$var2, numericSummaries = c("centralValue", "minMax"))
\end{Sinput}
\begin{Soutput}
     Feature         Result  
[1,] "Median"        "4.5"   
[2,] "Min. and max." "1; 999"
\end{Soutput}
\end{Schunk}

Note that the \code{summarize()}, \code{check()} and \code{visualize()} functions are also available interactively for full datasets, by calling e.g., \code{summarize(toyData)}. However, this produces an extensive amount of output in the console, and therefore, we generally do not recommend it, unless working with very small datasets or subsets of datasets.

@@

\section[Extending dataMaid]{Extending \pkg{dataMaid}}
\label{sec:extending}

\pkg{dataMaid} can be used as a user-friendly, self-contained package, as
shown in the previous sections. However, \pkg{dataMaid} is fully
customizable and \code{clean()} is mainly a tool for formatting the
results from various checking-, summary- and visualization
functions. Thus, the actual work underlying a \pkg{dataMaid} output file
can be anything --- depending on the arguments given to \code{clean()}
--- and user made functions are easily added to the
summarize/visualize/check-function arguments, as discussed previously.
However, the functions used in the SVC steps must adhere
to specific structures in order to be called from these three steps, and
therefore, we will now present how \code{summaryFunction}s,
\code{visualFunction}s and \code{checkFunction}s are made.

This section consists of two parts. First, we describe how customized \code{summaryFunction}s,
\code{visualFunction}s and \code{checkFunctions} can be made, including an introduction to a
few convenient tools available in \pkg{dataMaid} for aiding SVC-function construction.
Table \ref{table.functionTypes} serves as an overview of the internals of these three
central function types. Secondly, we turn to a
worked example of how to use custom made functions in practice. Here,
four new SVC functions are defined and used, both interactively and in \code{clean()}.

\begin{table}[htbp]
\footnotesize
\bgroup
\def\arraystretch{1.8}%
\begin{tabular}{L{0.2\linewidth}L{0.2\linewidth}L{0.2\linewidth}L{0.2\linewidth}}
& \code{summaryFunction} & \code{visualFunction} & \code{checkFunction} \\
\hline
\vspace{0pt} Input (required) & \code{v} - a variable vector \newline \code{...}  &  \code{v} - a variable vector \newline \code{vnam} - the variable name (as character string) \newline \code{doEval} - a logical (\code{TRUE}/\code{FALSE}) controlling the output type of the function & \code{v} - a variable vector \newline \code{nMax} - an integer (or \code{Inf}), controlling how many problematic values are printed, if relevant \newline \code{...}  \\
Input (optional) &  \code{maxDecimals} - number of decimals printed in outputted numerical values.  & - &  \code{maxDecimals}  - number of decimals printed in outputted numerical values.  \\
Purpose & Describe some aspect of the variable, e.g., a central value, its dispersion or level of missingness. & Produce a distribution plot. & Check a variable for a specific issue and, if relevant, identify the values in the variable that cause the issue. \\
Output (required) & A list with entries: \newline \code{\$feature} - a label for the summary value (as character string); \newline \code{\$result} - the result of the summary (as character string) & A character string with \code{R} code for producing a plot. This code should be standalone, i.e., should include the data if necessary. & A list with entries: \newline \code{\$problem} - a logical identifying whether an issue was found; \newline \code{\$message} - a character string (possibly empty) decribing the issue that was found, properly escaped and ready for use in \code{rmarkdown} \\
Output (recommended) & A \code{summaryResult} object, i.e., an attributed list with entries \code{\$feature}, \code{\$result} and \code{\$value}, the latter being the values from \code{\$result} in their original format). & \bgroup
\def\arraystretch{1.8}%
\hspace*{-.2cm}\begin{tabular}[t]{p{0.45\linewidth} |
                 p{0.45\linewidth}} \vspace*{-1cm}\raggedright If
                 \code{doEval} is \code{TRUE}: \newline A plot that is
                 opened by the graphic device in \code{R}. &
                                                          \vspace*{-1cm}\raggedright If \code{doEval} is \code{FALSE}: \newline   A text string with \code{R} code, as described above. \end{tabular} \egroup  & A \code{checkResult} object, i.e., an attributed list with entries \code{\$problem}, \code{\$message} and \code{\$problemValues}, the latter being either \code{NULL} or the problem causing values, as they were found in \code{v}, whichever is relevant. \\
Tools available for producing the function & \code{summaryResult()} & --- & \code{messageGenerator()} \newline \code{checkResult()}
\end{tabular}
\egroup
\caption{Reference information for creating new functions to be used
  as part of the \mbox{summarize-,} visualize-, and check steps. The three
  columns correspond to each of the three function types. %\hl{I think the formatting here is still sort of chaotic. Maybe include horizontal lines everywhere? Is it allowed in JSS? It's impossible to read what's in which row right now.}
}
\label{table.functionTypes}
\end{table}

\subsection{Function templates}
\label{sec:functionTemplates}
In order to construct a summary-, visual or check function, one needs
to create a new function with a specific structure.  This can be done with
different levels of strictness. If
the new custom function is only to be used as part of the SVC steps in \code{clean()},
then only the input/output structure of the function needs special attention.
However, new user-defined functions can also be registered locally to
be part of the full machinery of \pkg{dataMaid}, and these function will
be recognized and behave in the same way as the build-in functions in
\pkg{dataMaid}. The presentation below is given in the format of
function templates, written in pseudo-code. These templates are
designed for getting the full functionality, but note that
Table~\ref{table.functionTypes} serves as a reference to the minimal
requirements, while also presenting the "full" versions of the
function types.

@@

\subsubsection{Writing a summaryFunction}
As mentioned above, \pkg{dataMaid} provides a dedicated class for
\code{summaryFunction}s. However, this does not imply that they are
particularly advanced or complicated to create; in fact, they are
nothing but regular functions with a particular
input/output-structure. Specifically, they all follow the template
below:
\begin{Verbatim}
mySummaryFunction <- function(v, ...) {
  res <- [result of whatever summary we are doing]
  summaryResult(list(feature = "[Feature name]", result = res))
}
\end{Verbatim}
The last function called here, \code{summaryResult()}, changes the class
of the output, thereby making a \code{print()} method available for it.
Note that \code{v} is a vector and that \code{res} should be either a
character string or something that will be printed as one. In other
words, e.g., integers are allowed, but matrices are not. Though a lot
of different things can go into the \code{summaryFunction} template,
we recommend only using it for summarizing the features of a variable,
and leaving tests and checks for the \code{checkFunction}s (presented
below).

Adhering to the template above is sufficient for using the freshly
made \code{mySummaryFunction()} in \code{clean()}, but we recommend
furthermore adding the new function to the overview of all summary
functions by converting it to a proper \code{summaryFunction}
object. This is done by calling the \code{summaryFunction()} creator with
the user-defined function as the first argument, and additional arguments
\code{description} (an explanatory text which will be added to the attributes of
the function), and \code{classes} (a vector of variable classes the
user-defined function is intended to be applied to, also stored as an
attribute). In other words, a call on the following form should be made:
\begin{Schunk}
\begin{Sinput}
mySummaryFunction <- summaryFunction(mySummaryFunction,
  description = "[Text describing what the summaryFunction does]",
  classes = c([vector of data types that the function is intended for]))
\end{Sinput}
\end{Schunk}
which adds the new function to the output of an
\code{allSummaryFunctions()} call.
% One comment should be devoted to
% the two attributes of a \code{summaryFunction}.
If the \code{description} argument is left unspecified, the name of the
function (which in this case is \code{"mySummaryFunction"}) will be filled in and
stored under the \code{description} attribute. What happens if the \code{classes} argument is
not specified depends on the type of \code{mySummaryFunction}. If
\code{mySummaryFunction} is constructed as an \code{S3} generic function with associated
methods, the call to \code{summaryFunction()} will automatically produce
a vector of the names of the classes for which the function can be
called. If \code{mySummaryFunction} is not an \code{S3} generic and
\code{classes} is left unspecified, the attribute will simply be
left empty. Note that the helper function \code{allClasses()} might be useful
for filling out the \code{classes} argument, as it simply lists all
available classes in \pkg{dataMaid}:

\begin{Schunk}
\begin{Sinput}
R> allClasses()
\end{Sinput}
\begin{Soutput}
[1] "character" "Date"      "factor"    "integer"   "labelled"  "logical"  
[7] "numeric"  
\end{Soutput}
\end{Schunk}

% \hl{Write something here, don't end paragraph with code. Also, maybe move the allClasses() stuff somewhere else, it doesn't really belong under this header. Not sure where to, though.}

%Jeg foreslår at droppe nedenstående, da jeg synes det bliver lidt søgt i forhold
%til at det kun er beskrevet i pseudokode. (fx. er classes = numeric meget arbitrært)
%If we run the \code{allSummaryFunctions()} function we can see that the
%new function has been registered (output slightly appreviated):

@@

\subsubsection{Writing a visualFunction}
\code{visualFunction}s are the functions that produce the figures in a
\pkg{dataMaid} output document. Writing a \code{visualFunction} is slightly
more complicated than writing a \code{summaryFunction}. This follows from
the fact that \code{visualFunctions} need to be able to output standalone
code for plots in order for \code{clean()} to build standalone
\pkg{rmarkdown} files. We recommend using the following structure:
\begin{Schunk}
\begin{Sinput}
myVisualFunction <- function(v, vnam, doEval) {
  thisCall <- call("[the name of the function used to produce the plot]",
    v, [additional arguments for the plotting function])
  if (doEval) {
    return(eval(thisCall))
  } else return(deparse(thisCall))
}
\end{Sinput}
\end{Schunk}

In this function, \code{v} is the variable to be visualized,
\code{vnam} is its name (which should generally be passed to
\code{title} or \code{main} arguments in plotting functions) and
\code{doEval} controls whether the output is a plot (if \code{TRUE})
or a character string of standalone code for producing a plot (if
\code{FALSE}). Implementing the \code{doEval = TRUE} setting is not
strictly necessary for a \code{visualFunction}'s use in \code{clean()}, but
it makes it easier to assess what visualization options are available,
and obviously, it is crucial for interactive usage of
\code{myVisualFunction()}. In either case, it should be noted that all
the parameters listed above, \code{v}, \code{vnam} and \code{doEval},
are mandatory, so they must be left as is, even if they are not in
use (c.f. Table~\ref{table.functionTypes}).

As with the summary function, we call \code{visualFunction()} to
register our newly created function:
%. Like before, it accepts the
%user-defined function as the first argument, and two additional
%arguments description which adds explanatory flavour text as an
%attribute of the function, and a vector which determines which classes
%the user-defined function should be applied to (which is also stored
%as an attribute).
\begin{Schunk}
\begin{Sinput}
myVisualFunction <- visualFunction(myVisualFunction,
  description = "[Some text describing the visualFunction]",
  classes = c([data types that this function is intended for]))
  )
\end{Sinput}
\end{Schunk}

Now, \code{myVisualFunction()} will be available in a \code{allVisualFunctions()}
call, just like the two build-in \code{visualFunction}s, \code{standardVisual}
and \code{basicVisual}.
